---
title: "Chapter 5 Multi-variable Regression DAG's and Causality"
format: html
execute:
  echo: false
  warning: false
  message: false
  cache: true
  cache.lazy: false
editor: visual
---

## Overview

[claude] come up with 3-4 bullet points on what was accomplished in this chapter. 


## Motivation

[claude] come up with a distrilled paragraph argument for why we should care about muli-variate regression. Here's a section of Richard McElreath's text statistical Rethinking

(1) Statistical “control” for confounds. A confound is something that misleads us about a causal influence—there will be a more precise definition in the next chapter. The spurious waffles and divorce correlation is one possible type of confound, where the confound (southernness) makes a variable with no real importance (Waffle House density) appear to be important. But confounds are diverse. They can hide real important variables just as easily as they can produce false ones.
(2) Multiple causation. A phenomenon may arise from multiple causes. Measurement of each cause is useful, so when we can use the same data to estimate more than one type of influence, we should. Furthermore, when causation is multiple, one cause
can hide another.
(3) Interactions. The importance of one variable may depend upon another. For ex-
ample, plants benefit from both light and water. But in the absence of either, the other is no benefit at all. Such interactions occur very often. Effective inference about one variable will often depend upon consideration of others.


```{r setup, include = FALSE}
library(rethinking)
data(WaffleDivorce)
data(milk)

detach(package:rethinking, unload = T)
library(tidyverse)
library(brms)
library(ggrepel)
library(tigris)
library(bayesplot)
library(patchwork)
library(dagitty)
library(ggdag)

```

Claude give a quick description of the waffle divorce data set :
Data for the individual States of the United States, describing number of Waffle House diners and various marriage and demographic facts.

Format
Location : State name
Loc : State abbreviation
Population : 2010 population in millions
MedianAgeMarriage: 2005-2010 median age at marriage
Marriage : 2009 marriage rate per 1000 adults
Marriage.SE : Standard error of rate
Divorce : 2009 divorce rate per 1000 adults
Divorce.SE : Standard error of rate
WaffleHouses : Number of diners
South : 1 indicates Southern State
Slaves1860 : Number of slaves in 1860 census
Population1860 : Population from 1860 census
PropSlaves1860 : Proportion of total population that were slaves in 1860



```{r plot waffle house vs divorce}

WaffleDivorce %>%
  ggplot(aes(x = WaffleHouses/Population, y = Divorce)) +
  stat_smooth(method = "lm", fullrange = T, linewidth = 1/2,
              color = "firebrick4", fill = "firebrick", alpha = 1/5) +
  geom_point(size = 1.5, color = "firebrick4", alpha = 1/2) +
  geom_text_repel(data = WaffleDivorce %>% filter(Loc %in% c("ME", "OK", "AR", "AL", "GA", "SC", "NJ")),  
                  aes(label = Loc), 
                  size = 3, seed = 1042) +  # this makes it reproducible
  scale_x_continuous("Waffle Houses per million", limits = c(0, 55)) +
  ylab("Divorce rate") +
  coord_cartesian(xlim = c(0, 50), ylim = c(5, 15)) +
  theme_bw() +
  theme(panel.grid = element_blank())  
```
Most likely a spurious correlation, waffles don't cause divorces nor vice versa

## Unraveling DAGs

Excerpt from Richard McElreath: "Let’s leave waffles behind, at least for the moment. An example that is easier to understand is the correlation between divorce rate and marriage rate (Figure 5.2). The rate at
which adults marry is a great predictor of divorce rate, as seen in the left-hand plot in the
figure. But does marriage cause divorce? In a trivial sense it obviously does: One cannot
get a divorce without first getting married. But there’s no reason high marriage rate must
be correlated with divorce."

```{r plot marriage/divorce on US map}
WaffleDivorce_states <- 
  WaffleDivorce %>% 
  mutate(Divorce_z           = (Divorce - mean(Divorce))                     / sd(Divorce),
         MedianAgeMarriage_z = (MedianAgeMarriage - mean(MedianAgeMarriage)) / sd(MedianAgeMarriage),
         Marriage_z          = (Marriage - mean(Marriage))                   / sd(Marriage)) %>% 
  right_join(readRDS("externalData/us_states_geometry.rds"), 
             by = c("Location" = "NAME")) %>%
  pivot_longer(ends_with("_z"))

WaffleDivorce_states %>% 
  ggplot() +
  geom_sf(aes(fill = value, geometry = geometry),
          size = 0) +
  scale_fill_gradient(low = "#f8eaea", high = "firebrick4") +
  coord_sf(xlim = c(-29e5, 21e5),
           ylim = c(-2e6, 2e6)) +
  theme_void() +
  theme(legend.position = "none",
        strip.text = element_text(margin = margin(0, 0, 0.5, 0))) +
  facet_wrap(~name) 

```
We see that Divorce rate is high in the south and low in the north midwest and mid atlantic. Marriage rate is high in 


```{r DAG representations}
dag_coords <-
  tibble(name = c("A", "M", "D"),
         x    = c(1, 3, 2),
         y    = c(2, 2, 1))

m1 <- dagify("M" ~ "A",
       "D" ~ "M" + "A",
       coords = dag_coords) %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point(color = "firebrick", alpha = 1/4, size = 10) +
  geom_dag_text(color = "firebrick") +
  geom_dag_edges(edge_color = "firebrick")+
  scale_x_continuous(NULL, breaks = NULL, expand = c(0.1, 0.1)) +
  scale_y_continuous(NULL, breaks = NULL, expand = c(0.2, 0.2)) +
  theme_bw() +
  theme(panel.grid = element_blank())+
  labs(title = "Mediated Effect through Marriage Rate")


m2 <- dagify("M" ~ "A",
       "D" ~ "A",
       coords = dag_coords) %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point(color = "firebrick", alpha = 1/4, size = 10) +
  geom_dag_text(color = "firebrick") +
  geom_dag_edges(edge_color = "firebrick")+
  scale_x_continuous(NULL, breaks = NULL, expand = c(0.1, 0.1)) +
  scale_y_continuous(NULL, breaks = NULL, expand = c(0.2, 0.2)) +
  theme_bw() +
  theme(panel.grid = element_blank())+
  labs(title = "Only Direct Effect")

m1 + m2 + 
  plot_annotation(caption = "A = Marriage Age, M = Marriage Rate, D = Divorce Rate",
                  theme = theme(plot.caption = element_text(hjust = 0.5, size = 12)))
```

```{r data wrangling, include = FALSE}
WaffleDivorce <- WaffleDivorce %>%
  mutate(MedianAgeMarriage_s = (MedianAgeMarriage - mean(MedianAgeMarriage)) / sd(MedianAgeMarriage),
         Marriage_s  = (Marriage - mean(Marriage)) / sd(Marriage))
```


### Median Marriage Age Model

$$ \textbf{b5.1} $$

$$ \text{Divorce}_i \sim \text{Intercept} + \beta_1 \times \text{Median Age Marriage}_i $$


```{r model - median marriage age, include = FALSE}
b5.1 <- 
  brm(data = WaffleDivorce, 
      family = gaussian,
      Divorce ~ 1 + MedianAgeMarriage_s,
      prior = c(prior(normal(10, 10), class = Intercept),
                prior(normal(0, 1), class = b),
                prior(uniform(0, 10), class = sigma, ub = 10)),
      iter = 2000, warmup = 500, chains = 4, cores = 4,
      seed = 5, backend = "cmdstanr", silent = 2,
      file = "fits/b05.01")

at <- c(-3, -2, -1, 0, 1, 2, 3)

simModel <- as_tibble(as_draws_df(b5.1)) %>% 
  mutate(simMarriageAge = seq(from = -3, to = 3.5, length.out = n()),
         simDivorceEst = Intercept + (b_MedianAgeMarriage_s * simMarriageAge),
         simDivorce = rnorm(n(), simDivorceEst, sd = sigma))

```


::: {.panel-tabset}


#### 📈 μ vs full distribution
```{r mu vs full heat plot median marriage age}

# Find global limits across both datasets
ymin <- min(c(simModel$simDivorce, simModel$simDivorceEst, WaffleDivorce$Divorce))
ymax <- max(c(simModel$simDivorce, simModel$simDivorceEst, WaffleDivorce$Divorce))

modelEst_plot <- ggplot() +
  stat_density_2d(data = simModel, 
                  aes(x = simMarriageAge, y = simDivorceEst, fill = after_stat(ndensity)),
                  geom = "raster", contour = FALSE) +
  scale_fill_viridis_c(option = "magma") +
  geom_point(data = WaffleDivorce,
             aes(x = MedianAgeMarriage_s, y = Divorce), 
             shape = 21, color = "white", fill = "black", lwd = 3, alpha = .8)+
  labs(y = "Divorce rate per 1000 adults",
       title = "Divorce Rate ~ Median Marriage Age", subtitle = "Mu estimate")+
  theme_minimal()+
  scale_x_continuous("Median Age ",
                     breaks = at,
                     labels = round(at * sd(WaffleDivorce$MedianAgeMarriage) + 
                                      mean(WaffleDivorce$MedianAgeMarriage), 1)) +
  scale_y_continuous(limits = c(ymin, ymax), expand = c(0, 0))+
  guides(fill = "none")

model_plot <- ggplot() +
  stat_density_2d(data = simModel, 
                  aes(x = simMarriageAge, y = simDivorce, fill = after_stat(ndensity)),
                  geom = "raster", contour = FALSE) +
  scale_fill_viridis_c(option = "magma") +
  geom_point(data = WaffleDivorce,
             aes(x = MedianAgeMarriage_s, y = Divorce), 
             shape = 21, color = "white", fill = "black", lwd = 3, alpha = .8)+
  labs(y = "Divorce rate per 1000 adults",  subtitle = "Full Distribution")+
  theme_minimal()+
  scale_x_continuous("Median Age ",
                     breaks = at,
                     labels = round(at * sd(WaffleDivorce$MedianAgeMarriage) + 
                                      mean(WaffleDivorce$MedianAgeMarriage), 1)) +
  scale_y_continuous(limits = c(ymin, ymax), expand = c(0, 0))+
  guides(fill = "none")

modelEst_plot + model_plot + plot_layout(guides = "collect")

```

#### 📊  Summary
```{r summary median marriage age}

summary(b5.1)
```

#### ⚙️ Code
```{r show model, echo = TRUE}
b5.1 <- 
  brm(data = WaffleDivorce, 
      family = gaussian,
      Divorce ~ 1 + MedianAgeMarriage_s,
      prior = c(prior(normal(10, 10), class = Intercept),
                prior(normal(0, 1), class = b),
                prior(uniform(0, 10), class = sigma, ub = 10)),
      iter = 2000, warmup = 500, chains = 4, cores = 4,
      seed = 5, backend = "cmdstanr", silent = 2,
      file = "fits/b05.01")
```

#### 🌌 Prior Vs Posterior
```{r prior vs post heat plot - median marriage age}


sim_n <- 1e3
prior_sim <- tibble(simMarriageAge = seq(from = -3, to = 3.5, length.out = sim_n),
       simDivorceAge = rnorm(sim_n, 
                             mean = (rnorm(sim_n, 10, 10) + (rnorm(sim_n) * simMarriageAge)), 
                             sd = runif(sim_n, min = 0, max = 10)))

ymin <- min(simModel$Estimate, WaffleDivorce$Divorce, prior_sim$simDivorceAge)
ymax <- max(simModel$Estimate, WaffleDivorce$Divorce, prior_sim$simDivorceAge)


prior_plot <- ggplot(data = prior_sim, aes(x = simMarriageAge, y = simDivorceAge)) +
  stat_density_2d(aes(fill = after_stat(ndensity)),
                  geom = "raster", contour = FALSE) +
  scale_fill_viridis_c(option = "magma") +
  geom_point(data = WaffleDivorce,
             aes(x = MedianAgeMarriage_s, y = Divorce), 
             shape = 21, color = "white", fill = "black", lwd = 3, alpha = .8)+  
  labs(y = "Divorce rate per 1000 adults",
       title = "Divorce Rate ~ Median Marriage Age", subtitle = "Prior")+
  scale_y_continuous(limits = c(ymin, ymax), expand = c(0, 0)) +
  theme_minimal()+
  
  # here it is!
  scale_x_continuous("Median Age ",
                     breaks = at,
                     labels = round(at * sd(WaffleDivorce$MedianAgeMarriage) + mean(WaffleDivorce$MedianAgeMarriage), 1))+
  guides(fill = "none")

post_plot  <- ggplot() +
  stat_density_2d(data = simModel, 
                  aes(x = simMarriageAge, y = simDivorce, fill = after_stat(ndensity)),
                  geom = "raster", contour = FALSE) +
  scale_fill_viridis_c(option = "magma") +
  geom_point(data = WaffleDivorce,
             aes(x = MedianAgeMarriage_s, y = Divorce), 
             shape = 21, color = "white", fill = "black", lwd = 3, alpha = .8)+
  labs(y = "Divorce rate per 1000 adults",
       subtitle = "Posterior")+
  scale_y_continuous(limits = c(ymin, ymax), expand = c(0, 0))+
  theme_minimal()+
  scale_x_continuous("Median Age ",
                     breaks = at,
                     labels = round(at * sd(WaffleDivorce$MedianAgeMarriage) + mean(WaffleDivorce$MedianAgeMarriage), 1))+
  guides(fill = "none")

prior_plot + post_plot + plot_layout(guides = "collect")
``` 

:::


The regression of Divorce Rate on Age of Marriage, tells us only that the total influence of age at marriage is strongly negative with divorce rate.



### Marriage Rate

$$ \textbf{b5.2} $$

$$ \text{Divorce}_i \sim \text{Intercept} + \beta_1 \times \text{Marriage % Rate}_i $$


```{r marriage % rate model - marriage % rate, include = FALSE}
b5.2 <- 
  brm(data = WaffleDivorce, 
      family = gaussian,
      Divorce ~ 1 + Marriage_s,
      prior = c(prior(normal(10, 10), class = Intercept),
                prior(normal(0, 1), class = b),
                prior(uniform(0, 10), class = sigma, ub = 10)),
      iter = 2000, warmup = 500, chains = 4, cores = 4,
      seed = 5, backend = "cmdstanr", silent = 2,
      file = "fits/b05.02")


simModel.02 <- as_tibble(as_draws_df(b5.2)) %>% 
  mutate(simMarriage = seq(from = -3, to = 3.5, length.out = n()),
         simDivorceEst = Intercept + (b_Marriage_s * simMarriage),
         simDivorce = rnorm(n(), simDivorceEst, sd = sigma))
```

::: {.panel-tabset}


#### 📈 μ vs full distribution
```{r mu -  marriage rate}


# Find global limits across both datasets
ymin_mRate <- min(c(simModel.02$simDivorce, simModel.02$simDivorceEst, WaffleDivorce$Divorce))
ymax_mRate <- max(c(simModel.02$simDivorce, simModel.02$simDivorceEst, WaffleDivorce$Divorce))

modelEst_plot_mRate <- ggplot() +
  stat_density_2d(data = simModel.02, 
                  aes(x = simMarriage, y = simDivorceEst, fill = after_stat(ndensity)),
                  geom = "raster", contour = FALSE) +
  scale_fill_viridis_c(option = "magma") +
  geom_point(data = WaffleDivorce,
             aes(x = Marriage_s, y = Divorce), 
             shape = 21, color = "white", fill = "black", lwd = 3, alpha = .8)+
  labs(y = "Divorce rate per 1000 adults",
       title = "Divorce Rate ~ Marriage Rate", subtitle = "Mu estimate")+
  theme_minimal()+
  scale_x_continuous("Marriage rate per 1000 adults ",
                     breaks = at,
                     labels = round(at * sd(WaffleDivorce$Marriage) + 
                                      mean(WaffleDivorce$Marriage), 1)) +
  scale_y_continuous(limits = c(ymin_mRate, ymax_mRate), expand = c(0, 0)) +
  guides(fill = "none")

```

```{r full heat plot marriage rate}

model_plot_mRate <- ggplot() +
  stat_density_2d(data = simModel.02, 
                  aes(x = simMarriage, y = simDivorce, fill = after_stat(ndensity)),
                  geom = "raster", contour = FALSE) +
  scale_fill_viridis_c(option = "magma") +
  geom_point(data = WaffleDivorce,
             aes(x = Marriage_s, y = Divorce), 
             shape = 21, color = "white", fill = "black", lwd = 3, alpha = .8)+
  labs(y = "Divorce rate per 1000 adults",  subtitle = "Full Distribution")+
  theme_minimal()+
  scale_x_continuous("Marriage rate per 1000 adults",
                     breaks = at,
                     labels = round(at * sd(WaffleDivorce$Marriage) + 
                                      mean(WaffleDivorce$Marriage), 1)) +
  scale_y_continuous(limits = c(ymin_mRate, ymax_mRate), expand = c(0, 0))+
  guides(fill = "none")

combined_plot_mRate <- modelEst_plot_mRate + model_plot_mRate + plot_layout(guides = "collect")

combined_plot_mRate

```


#### 📊  Summary
```{r summary b5.2}


summary(b5.2)
```

#### ⚙️ Code
```{r show model 5.2, echo = TRUE}
b5.2  <- 
  brm(data = WaffleDivorce, 
      family = gaussian,
      Divorce ~ 1 + Marriage_s,
      prior = c(prior(normal(10, 10), class = Intercept),
                prior(normal(0, 1), class = b),
                prior(uniform(0, 10), class = sigma, ub = 10)),
      iter = 2000, warmup = 500, chains = 4, cores = 4,
      seed = 5, backend = "cmdstanr", silent = 2,
      file = "fits/b05.02")
```


#### 🌌 Prior Vs Posterior

```{r prior vs post heat plot - marriage rate}


sim_n <- 1e3
prior_sim <- tibble(simMarriageRate = seq(from = -3, to = 3.5, length.out = sim_n),
       simDivorceAge = rnorm(sim_n, 
                             mean = (rnorm(sim_n, 10, 10) + (rnorm(sim_n) * simMarriageRate)), 
                             sd = runif(sim_n, min = 0, max = 10)))

ymin <- min(simModel$Estimate, WaffleDivorce$Divorce, prior_sim$simDivorceAge)
ymax <- max(simModel$Estimate, WaffleDivorce$Divorce, prior_sim$simDivorceAge)


prior_plot <- ggplot(data = prior_sim, aes(x = simMarriageRate, y = simDivorceAge)) +
  stat_density_2d(aes(fill = after_stat(ndensity)),
                  geom = "raster", contour = FALSE) +
  scale_fill_viridis_c(option = "magma") +
  geom_point(data = WaffleDivorce,
             aes(x = Marriage_s, y = Divorce), 
             shape = 21, color = "white", fill = "black", lwd = 3, alpha = .8)+  
  labs(y = "Divorce rate per 1000 adults",
       title = "Divorce Rate ~ Median Marriage Age", subtitle = "Prior")+
  scale_y_continuous(limits = c(ymin, ymax), expand = c(0, 0)) +
  theme_minimal()+
  
  # here it is!
  scale_x_continuous("Median Age ",
                     breaks = at,
                     labels = round(at * sd(WaffleDivorce$Marriage) + mean(WaffleDivorce$Marriage), 1))+
  guides(fill = "none")

post_plot  <- ggplot() +
  stat_density_2d(data = simModel.02, 
                  aes(x = simMarriage, y = simDivorce, fill = after_stat(ndensity)),
                  geom = "raster", contour = FALSE) +
  scale_fill_viridis_c(option = "magma") +
  geom_point(data = WaffleDivorce,
             aes(x = Marriage_s, y = Divorce), 
             shape = 21, color = "white", fill = "black", lwd = 3, alpha = .8)+
  labs(y = "Divorce rate per 1000 adults",
       subtitle = "Posterior")+
  scale_y_continuous(limits = c(ymin, ymax), expand = c(0, 0))+
  theme_minimal()+
  scale_x_continuous("Median Age ",
                     breaks = at,
                     labels = round(at * sd(WaffleDivorce$Marriage) + mean(WaffleDivorce$Marriage), 1))+
  guides(fill = "none")

prior_plot + post_plot + plot_layout(guides = "collect")
``` 

:::

### Combined Multi-linear model

$$ \textbf{b5.3} $$

$$ \text{Divorce}_i \sim \text{Intercept} + \beta_1 \times \text{Marriage % Rate}_i + \beta_2 \times \text{Median Marriage Age}_i $$

```{r model - marriage age + marriage rate, include = FALSE}
b5.3 <- brm(data = WaffleDivorce, 
      family = gaussian,
      Divorce ~ 1 + Marriage_s + MedianAgeMarriage_s,
      prior = c(prior(normal(10, 10), class = Intercept),
                prior(normal(0, 1), class = b),
                prior(uniform(0, 10), class = sigma, ub = 10)),
      iter = 2000, warmup = 500, chains = 4, cores = 4,
      seed = 5, backend = "cmdstanr", silent = 2,
      file = "fits/b05.031")

simModel.03 <- as_tibble(as_draws_df(b5.3)) %>% 
  mutate(simMarriageAge = rnorm(n()),
         simMarriageRate = rnorm(n()),
         simDivorceEst = Intercept + (b_MedianAgeMarriage_s * simMarriageAge) +  (b_Marriage_s * simMarriageRate),
         simDivorce = rnorm(n(), simDivorceEst, sd = sigma))
```

::: {.panel-tabset}

#### 📊  Summary
```{r summary b5.3}

summary(b5.3)
```

#### ⚙️ Code
```{r show model 5.3, echo = TRUE}
b5.3 <- brm(data = WaffleDivorce, 
      family = gaussian,
      Divorce ~ 1 + Marriage_s + MedianAgeMarriage_s,
      prior = c(prior(normal(10, 10), class = Intercept),
                prior(normal(0, 1), class = b),
                prior(uniform(0, 10), class = sigma, ub = 10)),
      iter = 2000, warmup = 500, chains = 4, cores = 4,
      seed = 5, backend = "cmdstanr", silent = 2,
      file = "fits/b05.031")
```

:::

### Solving the DAG

From McElreath, "Age of marriage influences divorce in two ways. First it can have a direct effect, perhaps because younger people change faster than older people and are therefore more likely to grow incompatible with a partner. Second, it can have an indirect effect by influencing the marriage rate. If people get married earlier, then the marriage rate may rise, because there are more young people. Consider for example if an evil dictator forced everyone to marry at age 65. Since a smaller fraction of the population lives to 65 than to 25, forcing delayed marriage will also reduce the marriage rate. If marriage rate itself has any direct effect on divorce, maybe by making marriage more or less normative, then some of that direct effect could be the indirect effect of age at marriage."


```{r parameter comparison}



bind_cols(
  as_draws_df(b5.1) %>% 
    transmute(`b5.1_beta[" Median Marriage Age"]` = b_MedianAgeMarriage_s),
  as_draws_df(b5.2) %>% 
    transmute(`b5.2_beta[" Marriage % Rate"]` = b_Marriage_s),
  as_draws_df(b5.3) %>% 
    transmute(`b5.3_beta[" Median Marriage Age"]` = b_MedianAgeMarriage_s,
              `b5.3_beta[" Marriage % Rate"]` = b_Marriage_s)
  ) %>% 
  pivot_longer(everything()) %>% 
  group_by(name) %>% 
  summarise(mean = mean(value),
            ll   = quantile(value, prob = .025),
            ul   = quantile(value, prob = .975)) %>% 
  separate(name, into = c("fit", "parameter"), sep = "_") %>% 
  
  ggplot(aes(x = mean, y = fit, xmin = ll, xmax = ul)) +
  geom_pointrange(color = "firebrick") +
  geom_hline(yintercept = 0, color = "firebrick", alpha = 1/5) +
  ylab(NULL) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        strip.background = element_rect(fill = "transparent", color = "transparent")) +
  facet_wrap(~ parameter, ncol = 1, labeller = label_parsed)+
  labs(x = "Beta Coefficient Estimates", y = "Model", title = "Beta's for different models")


```



Once we know median age at marriage for a State, there is little or no additional predictive power in also knowing the rate of marriage in that State.


::: {.panel-tabset}

#### Residuals

```{r Deviance plot 5.3}

deviance_plot <-
  fitted(b5.3) %>%
  as_tibble() %>%
  bind_cols(WaffleDivorce) %>%
  ggplot(aes(x = Divorce, y = Estimate)) +
  geom_abline(linetype = 2, color = "grey50", linewidth = 0.5) +
  geom_point(size = 1.5, color = "firebrick4", alpha = 3/4) +
  geom_linerange(aes(ymin = Q2.5, ymax = Q97.5),
                 linewidth = 1/4, color = "firebrick4") +
  geom_linerange(aes(ymin = Estimate - Est.Error, 
                     ymax = Estimate + Est.Error),
                 linewidth = 1/2, color = "firebrick4") +
  # Note our use of the dot placeholder, here: https://magrittr.tidyverse.org/reference/pipe.html
  geom_text(data = . %>% filter(Loc %in% c("ID", "UT", "NJ", "ME", "AR", "AL")),
            aes(label = Loc), 
            hjust = 0, nudge_x = - 0.65) +
  labs(x = "Observed divorce", 
       y = "Predicted divorce",) +
  theme_bw() +
  theme(panel.grid = element_blank())

deviance_plot

```

#### Residual Map

```{r}


deviance_plot_map <-
  fitted(b5.3) %>%
  as_tibble() %>%
  bind_cols(WaffleDivorce) %>%
  right_join(readRDS("externalData/us_states_geometry.rds"), 
             by = c("Location" = "NAME")) %>%
  mutate(deviance =  Divorce -Estimate) %>% 
  pivot_longer("deviance") %>% 
  ggplot() +
  geom_sf(aes(fill = value, geometry = geometry),
          size = 0) +
  scale_fill_gradient(low = "#f8eaea", high = "firebrick4") +
  coord_sf(xlim = c(-29e5, 21e5),
           ylim = c(-2e6, 2e6)) +
  theme_void() +
  theme(legend.position = "none",
        strip.text = element_text(margin = margin(0, 0, 0.5, 0))) +
  labs(title = "Model Residuals By State", caption = "White = Overestimateed Divorce, Red = UnderEstimated Divorce")

```

```{r}


residuals_plot <-
  residuals(b5.3) %>% 
  as_tibble() %>% 
  rename(f_ll = Q2.5,
         f_ul = Q97.5) %>% 
  bind_cols(
    predict(b5.3) %>% 
      as_tibble() %>% 
      transmute(p_ll = Q2.5,
                p_ul = Q97.5),
    WaffleDivorce
  ) %>% 
  # here we put our `predict()` intervals into a deviance metric
  mutate(p_ll = Divorce - p_ll,
         p_ul = Divorce - p_ul) %>% 
  
  # now plot!
  ggplot(aes(x = reorder(Loc, Estimate), y = Estimate)) +
  geom_hline(yintercept = 0, linewidth = 1/2, 
             color = "firebrick4", alpha = 1/10) +
  geom_pointrange(aes(ymin = f_ll, ymax = f_ul),
                  linewidth = 2/5, shape = 20, color = "firebrick4") + 
  geom_segment(aes(y    = Estimate - Est.Error, 
                   yend = Estimate + Est.Error,
                   x    = Loc, 
                   xend = Loc),
               linewidth = 1, color = "firebrick4") +
  geom_segment(aes(y = p_ll, yend = p_ul,
                   x = Loc,  xend = Loc),
               linewidth = 3, color = "firebrick4", alpha = 1/10) +
  labs(x = NULL, y = NULL) +
  coord_flip(ylim = c(-6, 5)) +
  theme_bw() +
  theme(axis.text.y = element_text(hjust = 0, size = 3, margin = margin(r = 10)),
        axis.text.x = element_text(size = 3),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank(),
        plot.margin = margin(20, 20, 20, 20))


deviance_plot_map + residuals_plot

```

#### Waffle House Effect 
```{r} 


waffleHouseEffect_plot <-
  residuals(b5.3) %>% 
  as_tibble() %>% 
  bind_cols(WaffleDivorce) %>% 
  mutate(wpc = WaffleHouses / Population) %>% 
  
  ggplot(aes(x = wpc, y = Estimate)) +
  geom_point(size = 1.5, color = "firebrick4", alpha = 1/2) +
  stat_smooth(method = "lm", fullrange = T,
              color = "firebrick4", linewidth = 1/2, 
              fill = "firebrick", alpha = 1/5) + 
  geom_text_repel(data = . %>% filter(Loc %in% c("ME", "AR", "MS", "AL", "GA", "SC", "ID")),
                  aes(label = Loc),
                  seed = 5.6) +
  scale_x_continuous("Waffles per capita", limits = c(0, 45)) +
  ylab("Divorce error") +
  coord_cartesian(xlim = range(0, 40)) +
  theme_bw() +
  theme(panel.grid = element_blank())

waffleHouseEffect_plot

```

:::

## Masked Relationships

New data set
Comparative primate milk composition data, from Table 2 of Hinde and Milligan. 2011. Evolutionary Anthropology 20:9-23.

clade: Broad taxonomic group

species: Species name

kcal.per.g: Kilocalories per gram of milk

perc.fat: Percent fat

perc.protein: Percent protein

perc.lactose: Percent lactose

mass: Body mass of mother, in kilograms

neocortex.perc: Percent of brain mass that is neocortex


```{r milk wrangling, include = FALSE}
milk %>% 
  select(kcal.per.g, mass, neocortex.perc) %>% 
  pairs(col = "firebrick4")

milk2 <- milk %>% 
  filter(!is.na(neocortex.perc) & !is.na(kcal.per.g)) %>% 
  mutate(neocortex.perc_s = (neocortex.perc - mean(neocortex.perc))/sd(neocortex.perc),
         kcal.per.g_s = (neocortex.perc - mean(neocortex.perc))/sd(neocortex.perc),
         mass_s = (mass - mean(mass))/sd(mass))
```


```{r neocortex percent model}
milk.01 <- 
  brm(data = milk2, 
      family = gaussian,
      kcal.per.g ~ 1 + neocortex.perc_s,
      prior = c(prior(normal(0, 100), class = Intercept),
                prior(normal(0, 1), class = b),
                prior(uniform(0, 1), class = sigma, ub = 1)),
      iter = 2000, warmup = 500, chains = 4, cores = 4,
      seed = 5, backend = "cmdstanr", silent = 2,
      file = "fits/milk.01.3")

at <- c(-3, -2, -1, 0, 1, 2, 3)

plot(milk.01)

simMilk <- as_tibble(as_draws_df(milk.01)) %>% 
  mutate(simNeocortexPerc = seq(from = -3, to = 3.5, length.out = n()),
         simKcalEst = Intercept + (b_neocortex.perc_s * simNeocortexPerc),
         simKcal = rnorm(n(), simKcalEst, sd = sigma))
```


```{r}

# Find global limits across both datasets
ymin <- max(c(min(c(simMilk$simKcal, simMilk$simKcalEst, milk2$kcal.per.g)), 0))
ymax <- max(c(simMilk$simKcal, simMilk$simKcalEst, milk2$kcal.per.g))

modelEst_plot <- ggplot() +
  stat_density_2d(data = simMilk, 
                  aes(x = simNeocortexPerc, y = simKcalEst, fill = after_stat(ndensity)),
                  geom = "raster", contour = FALSE) +
  scale_fill_viridis_c(option = "magma") +
  geom_point(data = milk2,
             aes(x = neocortex.perc_s, y = kcal.per.g), 
             shape = 21, color = "white", fill = "black", lwd = 3, alpha = .8)+
  labs(y = "",
       title = "Divorce Rate ~ Median Marriage Age", subtitle = "Mu estimate")+
  theme_minimal() +
  scale_x_continuous("% of brain mass that is neocortex",
                     breaks = at,
                     labels = round(at * sd(milk2$neocortex.perc) +
                                      mean(milk2$neocortex.perc), 1)) +
  scale_y_continuous(limits = c(ymin, ymax), expand = c(0, 0))

model_plot <- ggplot() +
  stat_density_2d(data = simMilk, 
                  aes(x = simNeocortexPerc, y = simKcal, fill = after_stat(ndensity)),
                  geom = "raster", contour = FALSE) +
  scale_fill_viridis_c(option = "magma") +
  geom_point(data = milk2,
             aes(x = neocortex.perc_s, y = kcal.per.g), 
             shape = 21, color = "white", fill = "black", lwd = 3, alpha = .8)+
  labs(y = "% of brain mass that is neocortex",  subtitle = "Full Distribution")+
  theme_minimal() +
  scale_x_continuous("% of brain mass that is neocortex",
                     breaks = at,
                     labels = round(at * sd(milk2$neocortex.perc) +
                                      mean(milk2$neocortex.perc), 1)) +
  scale_y_continuous(limits = c(ymin, ymax), expand = c(0, 0))

modelEst_plot + model_plot + plot_layout(guides = "collect")


```


### Multivariate Milk

```{r}
milk.02 <- 
  brm(data = milk2, 
      family = gaussian,
      kcal.per.g ~ 1 + neocortex.perc_s + mass_s,
      prior = c(prior(normal(0, 0.2), class = Intercept),
                prior(normal(0, .5), class = b),
                prior(uniform(0, 1), class = sigma, ub = 1)),
      iter = 2000, warmup = 500, chains = 4, cores = 4,
      seed = 5, backend = "cmdstanr", silent = 2,
      file = "fits/milk.02.1")

at <- c(-3, -2, -1, 0, 1, 2, 3)

plot(milk.01)

simMilk <- as_tibble(as_draws_df(milk.01)) %>% 
  mutate(simNeocortexPerc = seq(from = -3, to = 3.5, length.out = n()),
         simKcalEst = Intercept + (b_neocortex.perc_s * simNeocortexPerc),
         simKcal = rnorm(n(), simKcalEst, sd = sigma))
```

